How to use replication assignments for teaching integrity in empirical archaeology

### 2019-06-23 analysis/paper/paper.Rmd
add edits from Hope

In his influential study of replication, sociologist of science Harry Collins argued that replication is at the core of scientific practice, writing “Replicability [...] is the Supreme Court of the scientific system” [@collins1992changing p. 19]. Like other observers of science, Collins claimed that in replication, private observations become communal facts, offering vital protection from error and fraud. In this paper we propose a new type of assignment for the archaeology classroom, the replication report, to better align the practice of teaching archaeology with the transparency and openness ideals of science [@nosek2015promoting]. The replication report assignment involves three key steps: 1) students analysing a published report to determine what are the main claims made by the authors of that report, 2) students obtaining the data used by the authors, and 3) analysing that data to determine if one or more of the authors’ claims are reliable.

We describe how to implement a replication report assignment suitable for upper level undergraduates and graduate students in archaeology. Our experience is based on an upper level archaeology class on stone artefact analysis taught during the Spring quarter of 2019 at the University of Washington. The class format includes a weekly cycle of lecture, discussion seminar, and hands-on laboratory activities. The assignments include seminar notes, lecture quizzes, laboratory worksheets, and two longer empirical reports. For the term that we report here, the class had 16 students, and one graduate student teaching assistant. This is a typical size for this class, and similar to the usual size of upper-level laboratory classes in the archaeology program at the University of Washington. Our students are mostly in social science and humanities majors with diverse levels of statistical knowledge. Here we survey the literature on similar types of assignments in other fields to identify common elements that other fields have identified as important principles and skills. We describe our assignment and discuss student feedback on our implementation. Finally, we offer recommendations for how to use replication reports to teach archaeology students.

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in @marwick2017open, we include all our assignment materials, as well as the entire R code used for all the analysis and visualizations contained in this paper, in our compendium at http://doi.org/10.17605/OSF.IO/DBSW9. Also in this version-controlled compendium are the raw data for the analyses reported here. The figures and results presented here can be independently reproduced with the code and data in this repository. In our compendium our code is released under the MIT licence, our data as CC-0, and our figures and assignment materials as CC-BY, to enable maximum re-use [for more details, see @Marwick2017].

@@

@barba2018terminologies points out that there is confusion and conflicting uses of the terms "reproducibility" and "replication" in recent years. In her survey of relevant literature from many disciplines, Barba finds that some fields make no distinction between "reproducibility" and "replication", and among fields that do recognise a distinction between the two, the meanings are sometimes directly swapped from one discipline to another. Here we follow what Barba has identified as the most common, long-established and highly cited definitions of these terms, and these are also the same as recently recommended by the National Academies of Sciences, Engineering, and Medicine (2019). Reproducibility is the ability to obtain results by using the same data, code, and procedures provided by the original authors [@Marwick2017]. This is only possible when the authors make all those materials available, for example in a research compendium [@marwick2018packaging]. Replication is the ability to arrive at the same scientific conclusions in a new study, collecting new data (possibly with different methods) and completing new analyses.

In the following section we briefly survey replication assignments described in other fields to survey the variety of forms this usually takes. Replication assignments across different fields may not strictly fit into the above definition of replication because they do not always involve a completely new study. We consider that if a study departs from any of the original materials e.g. new data with previously published code, or new code with previously published data, then it is broadly within the definition of replication.

@@

Students in Stanford University’s graduate course ‘Advanced Topics in Networking’ are given a replication assignment in which they are asked to replicating 'classic' computer networking experiments [@yan2017learning]. Students work in pairs and receive modest instructor support. The assignment entails selecting appropriate emulation software, communicating with the original authors, obtaining the authors’ materials, replicating the experiment, and publicising their results through both an in-class presentation and a blog post on a program website. We classify this as a replication assignment because many students could not obtain the original code from the authors, and had to write their own for the assignment. Since 2012, over 200 undergraduate and graduate students have participated in this assignment with an 86% success rate [@yan2017learning]. Student feedback suggests high satisfaction with the assignment, citing unique educational value, improved understanding of the original material, and the acquisition of professional skills. In some cases, students personally contributed to the network engineering literature when their replications exposed inaccuracies in original experiments, which were then presented to and publicly amended by the authors [@yan2017learning].

@@

For the first step we supplied students with a list of journal articles that included raw data and R code either in supplemental files or deposited on open data repositories. This list is updated regularly, but is not exhaustive, and is currently online at https://github.com/benmarwick/ctv-archaeology. Working in groups of 3-4 people, students selected a journal article from this list as their target article for this assignment. We encouraged them to choose a target article about a stone artefact analysis that looked interesting to them. We also required students to set up an open communication channel for their group to ensure they had an easy way to discuss their selection of the target article. We used Slack (https://slack.com/), a free cloud-based web application for team communication [@perkel2017scientists], to help them collaborate with each other efficiently. The instructors were members of all the student group channels to supervise, give guidance and support good communication habits. Students were required to individually submit the full bibliographic reference for their target article to complete step one.

@@

The second step was mostly straightforward, with students engaging in discussion in class and on Slack to identify the 2-3 key claims of their target paper, and identify the data visualisation that provided the most relevant support to one or more of those claims. Given that not all students have statistical background, the instructor covered some statistical methods they might encounter while reading during lecture, such as principal components analysis, to give them the mathematical concepts behind it. Identifying the data files was less straightforward, with about one third of students failing to correctly identify the data files accompanying their target article. We attribute this to the relatively low level of familiarity of the students in working with raw data, so they are not sure when they are looking at it, and the high degree of variability in how the target article authors make their data open. Some authors include their data as a file in the supplementary information attached to the article, while others deposit their files on an open data repository such as osf.io or figshare.com, and then cite the DOI to the files in their article. Where the data files were nested in several layers of folders, some students struggled to find them.

@@

We graded the students’ final submissions using a rubric with criteria that covered content, the introduction, methods, result, and conclusion sections, and style. In Figure \@ref(fig:show-plot-rubric) we show the distribution and means of student scores for each criterion. The two criteria showing the highest mean score are “Style: use commas and apostrophes correctly, and spell consistently”, and “Intro: has clear statement of the purpose of the report”. High scores for the grammar criterion are expected because these reflect basic writing skills required for many undergraduate-level courses. Students are expected to have learned these in lower level classes before taking this class. The high scores relating to the introduction section may reflect the effectiveness of the scaffolding steps that focused students on the specific purpose of the assignment. The lowest mean score is for “Content: minimum of 4 scholarly items in the reference list” that shows some people did not include four items. This might result from insufficient prior training in searching for scholarly publications, suggesting that although this is also a skill that should have been acquired before taking this class, many students remain weak at this task. A low mean is also evident for "Intro: has names, locations, and basic chronology of sites", because some students neglected to supply these archaeological details. Future use of this assignment will incorporate these low-scoring criteria into the scaffolding steps to emphasize their importance to students and give an opportunity for early feedback to students.

The criteria most relevant to the replication component of the assignment, in order from highest mean score to lowest, are "Content: submission includes Rmd file, Data file, and Word file",  "Conclusion: state whether the author’s claims appear to be robust, unreliable, etc", "Results: includes 1-2 original plots & description of these", and "Methods: identify the specific results you will replicate". This suggests that we could help students to develop a better skills in narrating their process (writing about methods), and describing and interpreting their data visualizations. In the future we may include more fundamental exercises focusing on these tasks in the scaffolding steps. Overall, we find that comparison of the scores for the replication criteria and other criteria shows there is no clear evidence that the replication component of this assignment lowers students’ grades. The two lowest scoring criteria are more generic research and writing skills rather than skills specific to the replication aspect of the assignment.

@@

Our main finding is that replication assignments are valued and in common use throughout the social sciences, and that they can be effective in teaching archaeology. Specifically, we found it possible to conduct a small scale replication assignment as part of an upper level undergraduate archaeology class. Student feedback indicated that it was a valuable new experience for them, and for the discipline, even if they could not see themselves doing it again. We found that although this was a new and unconventional assignment, these elements did not have a negative effect on students' grades. Although our study is limited by its small size, when considered with the numerous other reports of replication assignments in other fields, we believe this approach will work in many types of archaeology classes.  Replication assignments have an important role in closing the gap between the ideals of archaeological science and preparing students to tackle the practical challenges of doing archaeological science transparently and reproducibly.

To make it easier to conduct replication assignments with archaeology students in the future, we recommend instructors share their syllabi and assignment instructions in trustworthy repositories such as the Open Science Framework (https://osf.io/) or Dataverse (https://dataverse.harvard.edu/). Currently it is hard to find examples and a more systematic and open way of sharing might reduce preparation time for instructors [cf. @hoffler2017replicationwiki]. A second future direction in teaching replication is for archaeologists to share information about the software tools they use to make reproducible research easier. This information can be useful to guide instructors on what to teach students as part of their methods and software training [@janz2016bringing]. In our review above we noted that teaching the use of tools like R, Markdown, Git and GitHub has already been embraced in many fields as core elements of graduate programs. Archaeology programs must place a greater emphasis on giving students skills to use tools that enable transparent and reproducible research.

>>>>

In his influential study of replication, sociologist of science Harry Collins argued that replication is at the core of scientific practice, writing “Replicability [...] is the Supreme Court of the scientific system” [@collins1992changing p. 19]. Like other observers of science, Collins claimed that in replication, private observations become communal facts, offering vital protection from error and fraud. In this paper we propose a new type of assignment for the archaeology classroom, the replication report, to better align the practice of teaching archaeology with the transparency and openness ideals of science [@nosek2015promoting]. The replication report assignment involves three key steps: 1) students analysing a published report to determine the main claims made by the authors of that report, 2) students obtaining the data used by the authors, and 3) analysing that data to determine if one or more of the authors’ claims are reliable.

We describe how to implement a replication report assignment suitable for upper-level undergraduates and graduate students in archaeology. Our experience is based on an upper-level archaeology class on stone artefact analysis taught during the Spring quarter of 2019 at the University of Washington. The class format includes a weekly cycle of lecture, discussion seminar, and hands-on laboratory activities. The assignments include seminar notes, lecture quizzes, laboratory worksheets, and two longer empirical reports. For the term that we report here, the class had 16 students and one graduate student teaching assistant. This is a typical size for this class, and similar to the usual size of upper-level laboratory classes in the archaeology program at the University of Washington. Our students are mostly in social science and humanities majors with varying levels of statistical competence. Here we survey the literature on similar types of assignments in other fields to identify common elements that other fields have identified as important principles and skills. We describe our assignment and discuss student feedback on our implementation. Finally, we offer recommendations for how to use replication reports to teach archaeology students.

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in @marwick2017open, we include all our assignment materials, as well as the entire R code used for all the analysis and visualizations contained in this paper, in our compendium at http://doi.org/10.17605/OSF.IO/DBSW9. Also in this version-controlled compendium are the raw data for the analyses reported here. The figures and results presented here can be independently reproduced with the code and data in this repository. In our compendium our code is released under the MIT license, our data as CC-0, and our figures and assignment materials as CC-BY, to enable maximum re-use [for more details, see @Marwick2017].

@@

@barba2018terminologies points out that although there has been prolific discussion of the terms "reproducibility" and "replication" in many disciplines in recent years, confusion and conflicting uses are widespread. In her survey of relevant literature, Barba finds that some fields make no distinction between "reproducibility" and "replication", and among fields that do recognise a distinction between the two, the meanings are sometimes directly inverted. Here we follow what Barba has identified as the most common, long-established and highly cited definitions of these terms, as also recently recommended by the National Academies of Sciences, Engineering, and Medicine (2019). Reproducibility is the ability to obtain results by using the same data, code, and procedures provided by the original authors [@Marwick2017]. This is only possible when the authors make all those materials available, for example in a research compendium [@marwick2018packaging]. Replication is the ability to arrive at the same scientific conclusions in a new study, collecting new data (possibly with different methods) and completing new analyses.

In the following section we briefly survey replication assignments described in other fields to survey the variety of forms this usually takes. Replication assignments across different fields may not strictly fit into the above definition of replication because they do not always involve a completely new study. Nevertheless, we consider that if a study departs from any of the original materials, e.g. new data with previously published code, or new code with previously published data, then it is broadly within the definition of replication.

@@

Students in Stanford University’s graduate course ‘Advanced Topics in Networking’ are given a replication assignment in which they are asked to replicate 'classic' computer networking experiments [@yan2017learning]. Students work in pairs and receive modest instructor support. The assignment entails selecting appropriate emulation software, communicating with the original authors, obtaining the authors’ materials, replicating the experiment, and publicising their results through both an in-class presentation and a blog post on a program website. We classify this as a replication assignment because many students could not obtain the original code from the authors, and had to write their own for the assignment. Since 2012, over 200 undergraduate and graduate students have participated in this assignment with an 86% success rate [@yan2017learning]. Student feedback suggests high satisfaction with the assignment, citing unique educational value, improved understanding of the original material, and the acquisition of professional skills. In some cases, students personally contributed to the network engineering literature when their replications exposed inaccuracies in original experiments, which were then presented to and publicly amended by the authors [@yan2017learning].

@@

For the first step, we supplied students with a list of journal articles that included raw data and R code either in supplemental files or deposited on open data repositories. This list is updated regularly, but is not exhaustive, and is currently online at https://github.com/benmarwick/ctv-archaeology. Working in groups of 3-4 people, students selected a journal article from this list as their target article for this assignment. We encouraged them to choose a target article about a stone artefact analysis that looked interesting to them. We also required students to set up an open communication channel for their group to ensure they had an easy way to discuss their selection of the target article. We used Slack (https://slack.com/), a free cloud-based web application for team communication [@perkel2017scientists], to help them collaborate with each other efficiently. The instructors were members of all the student group channels to supervise, give guidance and support good communication habits. Students were required to individually submit the full bibliographic reference for their target article to complete step one.

@@

The second step was mostly straightforward, with students engaging in discussion in class and on Slack to identify the 2-3 key claims of their target paper, and identify the data visualisation that provided the most relevant support to one or more of those claims. Given that not all students have statistical background, the instructor covered some statistical methods they might encounter while reading during lecture, such as principal components analysis, to give them the mathematical concepts behind it. Identifying the data files was less straightforward, with about one third of students failing to correctly identify the data files accompanying their target article. We attribute this to the relatively low level of familiarity of the students in working with raw data, so they are not sure when they are looking at it, and the high degree of variability in how the target article authors make their data open. Some authors include their data as a file in the supplementary information attached to the article, while others deposit their files on an open data repository such as osf.io or figshare.com, and then cite the DOI to the files in their article. When the data files were nested in several layers of folders, some students struggled to find them.

@@

We graded the students’ final submissions using a rubric with criteria that covered content, the introduction, methods, result, and conclusion sections, and style. In Figure \@ref(fig:show-plot-rubric) we show the distribution and means of student scores for each criterion. The two criteria showing the highest mean score are “Style: use commas and apostrophes correctly, and spell consistently”, and “Intro: has clear statement of the purpose of the report”. High scores for the grammar criterion are expected because these reflect basic writing skills required for many undergraduate-level courses. Students are expected to have learned these in lower-level classes before taking this class. The high scores relating to the introduction section may reflect the effectiveness of the scaffolding steps that focused students on the specific purpose of the assignment. The lowest mean score is for “Content: minimum of 4 scholarly items in the reference list” that shows some people did not include four items. This might result from insufficient prior training in searching for scholarly publications, suggesting that although this is also a skill that should have been acquired before taking this class, many students remain weak at this task. A low mean is also evident for "Intro: has names, locations, and basic chronology of sites," because some students neglected to supply these archaeological details. Future use of this assignment will incorporate these low-scoring criteria into the scaffolding steps to emphasize their importance to students and give an opportunity for early feedback to students.

The criteria most relevant to the replication component of the assignment, in order from highest mean score to lowest, are "Content: submission includes Rmd file, Data file, and Word file",  "Conclusion: state whether the author’s claims appear to be robust, unreliable, etc", "Results: includes 1-2 original plots & description of these", and "Methods: identify the specific results you will replicate". This suggests that we could help students to develop better skills in narrating their process (writing about methods), and describing and interpreting their data visualizations. In the future we may include more fundamental exercises focusing on these tasks in the scaffolding steps. Overall, we find that comparison of the scores for the replication criteria and other criteria shows there is no clear evidence that the replication component of this assignment lowers students’ grades. The two lowest scoring criteria are more generic research and writing skills rather than skills specific to the replication aspect of the assignment.

@@

Our main finding is that replication assignments are valued and in common use throughout the social sciences, and that they can be effective in teaching archaeology. Specifically, we found it possible to conduct a small scale replication assignment as part of an upper-level undergraduate archaeology class. Student feedback indicated that it was a valuable new experience for them, and for the discipline, even if they could not see themselves doing it again. We found that although this was a new and unconventional assignment, these elements did not have a negative effect on students' grades. Although our study is limited by its small size, when considered with the numerous other reports of replication assignments in other fields, we believe this approach will work in many types of archaeology classes.  Replication assignments have an important role in closing the gap between the ideals of archaeological science and preparing students to tackle the practical challenges of doing archaeological science transparently and reproducibly.

To make it easier to conduct replication assignments with archaeology students in the future, we recommend instructors share their syllabi and assignment instructions in trustworthy repositories such as the Open Science Framework (https://osf.io/) or Dataverse (https://dataverse.harvard.edu/). Currently, it is hard to find examples and a more systematic and open way of sharing might reduce preparation time for instructors [cf. @hoffler2017replicationwiki]. A second future direction in teaching replication is for archaeologists to share information about the software tools they use to make reproducible research easier. This information can be useful to guide instructors on what to teach students as part of their methods and software training [@janz2016bringing]. In our review above we noted that teaching the use of tools like R, Markdown, Git and GitHub has already been embraced in many fields as core elements of graduate programs. Archaeology programs must place a greater emphasis on giving students skills to use tools that enable transparent and reproducible research.
